---
params: 
  continent: 
title: "`r paste(params$continent, 'Report')`"
author: Paul Campbell
description: "Create reproducible reports from data cleaning to visualisation with rmarkdown"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    theme: yeti
    highlight: haddock
---

```{r include=FALSE}
knitr::opts_chunk$set(fig.width = 9, 
                      fig.height = 6,
                      fig.align = "center", 
                      cache = TRUE,
                      cache.path = ".cache/",
                      message=FALSE, 
                      warning=FALSE, 
                      echo=FALSE
                      )
```

```{r setup, cache=FALSE}
library(tidyverse)
library(readxl)
library(hrbrthemes)
library(scales)
library(countrycode)
library(glue)
library(plotly)
library(crosstalk)
library(DT)
library(gganimate)
library(widgetframe)

theme_set(theme_ft_rc())
```

---

This is an example of a parameterised HTML report produced with `rmarkdown`. Given a dataset, an `rmarkdown` script, a list of parameters to pass to it (in this case Continents) and a rendering function, we can batch produce a report for each Continent supplied to the report script.

The data, script, render function and generated reports are viewable in this [github repository](https://github.com/PaulC91/gapminder_reports).

For information about automated data reporting wit R, check out this [blogpost on the Culture of Insight website](https://www.cultureofinsight.com/blog/2018/10/22/2018-08-20-automated-data-reports-with-r/).


```{r cache=TRUE}
path <- "gapminder_messy.xlsx"

combined_data <- 
  excel_sheets(path) %>% 
  map_df(~ {
    read_excel(path, sheet = .x, skip = 4, trim_ws = TRUE) %>% 
      mutate(year = as.integer(.x))
  }) %>% 
  select(country, year, everything())
```

```{r}
combined_data <- combined_data %>% 
  mutate(continent = countrycode(sourcevar = country, origin = "country.name", destination = "continent")) %>% 
  select(continent, everything())
```

```{r}
filtered_data <- combined_data %>% 
  filter(continent == params$continent)

latest_year <- max(filtered_data$year)
years <- unique(filtered_data$year)
n_year <- length(years)
previous_year <- sort(years, partial=n_year-1)[n_year-1]

```

## Linked Interactive Graphics

Chart showing latest percentage changes in metrics from year `r previous_year ` to `r latest_year`. 

Use the search bar to highlight a particular country across all 3 charts or click directly on one of the bars. Double clicking will deselect the highlighted country.

```{r}
p_change <- filtered_data %>% 
  group_by(continent, country) %>% 
  arrange(year) %>% 
  mutate(gdpPercap_change = (gdpPercap - lag(gdpPercap)) / lag(gdpPercap),
         lifeExp_change = (lifeExp - lag(lifeExp)) / lag(lifeExp),
         pop_change = (pop - lag(pop)) / lag(pop))
         
latest_figs <- p_change %>% 
  filter(year == max(year)) %>%
  select(-year) %>% 
  arrange(country) %>% 
  ungroup()

```

```{r linked-charts, eval=TRUE, echo=FALSE}
# create a shared data source with the crosstalk package
sd <- SharedData$new(latest_figs, ~country, group = "Highlight a Country")

# build 3 static ggplot charts with the shared data
sd_gdp <- ggplot(sd, aes(reorder(country, -gdpPercap_change), gdpPercap_change,
                               text = glue::glue("{country}, {continent}: {percent(gdpPercap_change)}"))) +
  geom_col(aes(fill = gdpPercap_change)) +
  scale_y_percent() +
  scale_fill_viridis_c(guide = FALSE) +
  labs(title = "GDP Per Capita % Change", y = "GDP per capita") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        panel.grid.major.x = element_blank())

sd_le <- ggplot(sd, aes(reorder(country, -lifeExp_change), lifeExp_change,
                               text = glue::glue("{country}, {continent}: {percent(lifeExp_change)}"))) +
  geom_col(aes(fill = lifeExp_change)) +
  scale_y_percent() +
  scale_fill_viridis_c(guide = FALSE) +
  labs(title = "Life Expectacy % Change", y = "Life Expectancy") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        panel.grid.major.x = element_blank())

sd_pop <- ggplot(sd, aes(reorder(country, -pop_change), pop_change,
                               text = glue::glue("{country}, {continent}: {percent(pop_change)}"))) +
  geom_col(aes(fill = pop_change)) +
  scale_y_percent() +
  scale_fill_viridis_c(guide = FALSE) +
  labs(title = "Latest Period % Change", y = "Population") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        panel.grid.major.x = element_blank())

# make them interactive with plotly
sd_gdp_i <- ggplotly(sd_gdp, tooltip = "text")
sd_le_i <- ggplotly(sd_le, tooltip = "text")
sd_pop_i <- ggplotly(sd_pop, tooltip = "text")

# arrange them together
subplot(sd_gdp_i, sd_le_i, sd_pop_i, nrows = 3, titleY = TRUE) %>% 
  config(collaborate = FALSE, displaylogo = FALSE) %>% 
  hide_legend() %>% 
  highlight(selectize = TRUE, on = "plotly_click", off = "plotly_doubleclick") %>% 
  frameWidget(height = "600px")

```

<br>

---

## Latest Period Figures (`r latest_year`) & % Changes (`r previous_year`-`r latest_year`)

Searchable + sortable data table.

```{r echo=FALSE}

mills <- scales::unit_format(scale = 1e-6, accuracy = .01, unit = "M")

tdat <- latest_figs %>%
  select(-continent) %>% 
  select(1:2, 5, 3, 6, 4, 7) %>% 
  mutate(pop = mills(pop))

DT::datatable(tdat, rownames = FALSE, 
             colnames = c("Country", "GDP Per Capita", "+/-", "Life Expectancy", "+/-", "Population", "+/-"),
             options = list(dom = 'ftip',
                            columnDefs = list(list(className = 'dt-center', targets = c(1,3,5))))
             ) %>% 
  formatCurrency(columns = 2) %>% 
  formatRound(columns = 4, digits = 1) %>% 
  formatPercentage(columns = c(3,5,7))
```

---

## Static Charts with ggplot2

Bubble chart visualising all 3 metrics for latest year of data with automated placement of annotations.

```{r echo=FALSE, dpi = 300, dev = "svg"}
filtered_data %>% 
  filter(year == max(year)) %>% 
  ggplot(aes(gdpPercap, lifeExp, label = country)) +
  geom_point(aes(size = pop), fill = "SteelBlue", colour = "white", alpha = .7, pch = 21) +
  ggrepel::geom_text_repel(colour = "white", force = 50, segment.colour = "grey", size = 3) +
  scale_size(range = c(2, 12)) +
  labs(x = "GDP per capita", y = "Life Expectancy",
       title = glue("Latest {params$continent} Metrics"), subtitle = glue("Year: {latest_year}"),
       caption = "@paulcampbell91 | Culture of Insight | Source: Gapminder", size = "Population")
```

---

## Animated Charts

```{r}
continent_summary <- 
  combined_data %>% 
  group_by(continent, year) %>% 
  summarise(gdpPercap = weighted.mean(gdpPercap, pop),
            lifeExp = weighted.mean(lifeExp, pop),
            pop = sum(pop)) %>% 
  ungroup() %>% 
  gather(metric, value, 3:5)
```

Add animation to the charts in your reports for greater impact. 

```{r echo=FALSE, fig.height=7, eval=TRUE}
ggplot(continent_summary, aes(year, value, group = continent)) +
  facet_wrap(~metric, ncol = 1, scales = "free_y") +
  geom_line(aes(colour = continent)) +
  geom_segment(aes(xend = 2007, yend = value), linetype = 2, colour = 'grey') + 
  geom_point(size = 2, colour = "white") + 
  geom_text(aes(x = 2007.1, label = continent), hjust = 0) + 
  transition_reveal(continent, year) + 
  coord_cartesian(clip = 'off') + 
  labs(title = "Continental Progress", y = NULL, 
       caption = "@paulcampbell91 | Culture of Insight | Source: Gapminder") +
  theme(plot.margin = margin(5.5, 40, 5.5, 5.5), legend.position = "none")
```

---

## Learning

Making the switch to doing your data work with R can be daunting, but with the combination of <a href="https://www.rstudio.com/products/RStudio/" target="_blank">RStudio</a> as your 'integrated development environment', the <a href="https://www.tidyverse.org/" target="_blank">tidyverse packages</a> for data analysis, and <a href="https://rmarkdown.rstudio.com/" target="_blank">rmarkdown</a> for producing reports in just about any format you'd like (HTML, PDF, Word, PowerPoint, Dashboard, even entire websites like this one!), there has never been a better time to make the leap and start your journey to better data reporting workflows. There's also a really great and welcoming community of R users online who are always happy to help new users feel at home.

If you'd like some help getting started, we're currently running a 1-day 'Master Data with Code' workshop where you'll learn how to import, manipulate, join and transform data using the `tidyverse` in the RStudio IDE.

We also offer bespoke in-house training for teams where we focus on your own specific data tasks and how you can use R for greater accuracy and efficiency in data processing, and engagement in your communicated insights.

For more information get in touch via our [Contact Page](https://cultureofinsight.com/contact/).

Thanks for reading!
